Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-12 large_number_squeezing_with_atoms.tex
Add and order refs, discuss the point of our analytic solution
Also various other scattered edits

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large for arbitrarily large condensates, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with particular trap geometries.  We also introduce a method of maximising the measurable squeezing by using a pi-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

Quantum degenerate ultracold atomic sources have been used to study fundamental processes such as quantum phase transitions \cite{Leggett2001,Buluta2009}, quantum non-equilibrium thermodynamics \cite{Kinoshita:2006,Barnett:2011} and entanglement of massive particles \cite{Kheruntsyan:2005}.  Their ability to be highly isolated from their environment and controlled precisely by optical, magnetic and rf fields makes them an ideal platform for engineering and manipulating non-trivial quantum states. In quantum optics, the most common way to produce entangled states has been through squeezing, and it has been recognised that the intrinsic nonlinearities of atoms naturally lead to quadrature squeezing \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, doeringET2010}.  Number squeezing can be achieved with spin squeezing techniques \cite{kitagawaET1993, gross2010, sinatra2012}, molecular dissociation of Bose-Einstein Condensates (BECs) \cite{kheruntsyanET2002, davisET2008, savage2007, ogren2010}, four-wave mixing \cite{perrin2008, ogren2009, jascula2010}, double-well potentials \cite{esteve2008, maussang2010}, or mixing quadrature-squeezed states using Ramsey interferometry \cite{haineET2009}. The creation of these non-classical atomic states has relevance for fundamental tests of quantum mechanics such as the EPR paradox \cite{pu2000, sorensen2001, haine2005,zhao2007, ogren2010}, as well as atom inteferometry \cite{sorensenET1999, kuzmichET2000, meyerET2001, liebfriedET2004, roosET2006, doeringET2010, gross2010}.
While these techniques have resulted in significant experimentally detectable number squeezing \cite{gross2010}, they have been limited to small numbers of atoms, typically a few hundred.

@@

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit, the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be effectively squeezed, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes.
Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{kheruntsyanET2002, Kheruntsyan:2005, johnssonET2007, haineET2009}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

In section \ref{secTwoModeAnalytic} we examine the ideal behaviour when the system is reduced to a single spatial mode, extracting a full analytic solution to the squeezing for arbitrary \textbf{whatever it is we extended, along with reference of what was covered previously.}  In section \ref{sec:MMdescription} we explore the idea of how multimode behaviour will become important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Section~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Section~\ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can dramatically improve mode-matching issues.  Section~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the  scaling with dimensionality.  Section~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

@@

The system is prepared with all the population in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to be initially a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As the system is insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states.  This state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

@@

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\theta) \label{eqsDef} \\
c &=& \cos(\theta) \\
\theta &=& \Omega(t_3 - t_2) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \nonumber\\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) \left( n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B \right. \nonumber \\
       && \,\,\,\,\, \left. + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^* \right) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{johnssonET2007,johnssonHope:2007,haineET2009,dennisET2010,steelET1998}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against both the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such as the appearance of negative densities in lightly populated modes.  We could also use the Bogoliubov analysis described in section \ref{sec:2D3Dsqueezing} as an independent estimate of the validity of the TWA.  We found that the TWA remained valid except in some 3D simulations with high nonlinearities, where there were a large number of modes, and the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown.  Fortunately, the best squeezing was typically found in regions where the TWA was valid.

@@

We have presented an analytic solution to the two-mode model of atomic fields with a strong Kerr non-linearity, allowing for a rapid estimate of the atomic number difference squeezing that can be obtained from these systems as scattering length data is obtained.  \textbf{Reiterate what is new in this analytic solution}.  In general, it predicts arbitrarily good squeezing in the two-mode limit. We then examined trap geometries in which this limit can be approached.

Using a constant potential in the trap (a `box' trap) allows the BEC density to be spatially uniform throughout the process.  This is shown to alleviate the difficulties with choosing correct hold times and phases for the coupling pulses, and also alleviates the majority of the mode-matching issues for the final recombination of the two modes. If such a box potential is not available, the application of a pi-pulse part way through the hold period can also improve mode matching by imparting the same modal dynamics to both states.

@@

\bibitem{Buluta2009} Buluta I and Nori F 2009 Quantum Simulators \emph{Science} {\bf 326} 108

@@

@@

\bibitem{jascula2010} Jaskula J C, Bonneau M, Partridge G B, Krachmalnicoff V, Deuar P, Kheruntsyan K V, Aspect A, Boiron D, and Westbrook C I 2010 Sub-Poissonian Number Differences in Four-Wave Mixing of Matter Waves \emph{Phys. Rev. Lett.} {\bf 105} 190402

@@

\bibitem{savage2007} Savage C M and Kheruntsyan K V 2007 Spatial Pair Correlations of Atoms in Molecular Dissociation \emph{Phys. Rev. Lett.} {\bf 99} 220404

\bibitem{pu2000} Pu H and Meystre P 2000 Creating Macroscopic Atomic Einstein-Podolsky-Rosen States from Bose-Einstein Condensates \emph{Phys. Rev. Lett.} {\bf 85} 3987

\bibitem{esteve2008} Est{\`{e}}eve J, Gross C, Weller A, Giovanazzi S, and Oberthaler M 2008 Squeezing and entanglement in a Bose-Einstein condensate \emph{Nature} {\bf 455} 1216

@@

@@

\bibitem{gaunt2013} Gaunt A L, Schmidutz T F, Gotlibovych I, Smith R P, and Hadzibabic Z 2013 Bose-Einstein Condensation of Atoms in a Uniform Potential \emph{Phys. Rev. Lett.} {\bf 110} 200406

\bibitem{cornish:2000} Cornish S, Claussen N, Roberts J, Cornell E, and Wieman C 2000 Stable $^{85}$Rb Bose-Einstein Condensates with Widely Tunable Interactions \emph{Phys. Rev. Lett.} {\bf 85} 1795

>>>>

\begin{abstract}
We examine the feasibility of creating and measuring large relative number squeezing in multicomponent trapped Bose-Einstein condensates.  In the absence of multimode effects, this squeezing can be arbitrarily large for arbitrarily large condensates, but a range of processes limit the measurable squeezing in realistic trap configurations.  We examine these processes, and suggest methods to mitigate them. We conclude that high levels of squeezing with large numbers of atoms is feasible, but can realistically only be achieved in traps with particular trap geometries.  We also introduce a method of maximising the measurable squeezing by using a $\pi$-pulse during the process to improve spatial mode-matching.
\end{abstract}

@@

Quantum degenerate ultracold atomic sources have been used to study fundamental processes such as quantum phase transitions \cite{Leggett2001,Buluta2009}, quantum non-equilibrium thermodynamics \cite{Kinoshita:2006,Barnett:2011} and entanglement of massive particles \cite{Kheruntsyan:2005}.  Their ability to be highly isolated from their environment and controlled precisely by optical, magnetic and rf fields makes them an ideal platform for engineering and manipulating non-trivial quantum states. In quantum optics, the most common way to produce entangled states has been through squeezing, and it has been recognised that the intrinsic nonlinearities of atoms naturally lead to quadrature squeezing \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, doeringET2010}.  Number squeezing can be achieved with spin squeezing techniques \cite{kitagawaET1993, gross2010, sinatra2012}, molecular dissociation of Bose-Einstein Condensates (BECs) \cite{kheruntsyanET2002,davisET2008,savage2007,ogren2010}, four-wave mixing \cite{perrin2008, ogren2009, jascula2010}, double-well potentials \cite{esteve2008, maussang2010}, or mixing quadrature-squeezed states using Ramsey interferometry \cite{haineET2009}. The creation of these non-classical atomic states has relevance for fundamental tests of quantum mechanics such as the EPR paradox \cite{ogren2010, pu2000, sorensen2001, haine2005,zhao2007}, as well as atom interferometry \cite{doeringET2010, gross2010, sorensenET1999, kuzmichET2000, meyerET2001, liebfriedET2004, roosET2006}.
While these techniques have resulted in significant experimentally detectable number squeezing \cite{esteve2008,gross2010}, they have been limited to small numbers of atoms, typically a few hundred.

@@

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit, the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be effectively squeezed, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes.
Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, kheruntsyanET2002}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

In Section \ref{secTwoModeAnalytic} we examine the ideal behaviour of a nonlinear bosonic two-state system when that system possesses only a single spatial mode, extracting full analytic solutions for the number squeezing in each of the two states, as well as the number difference squeezing between the two states. These solutions of the ideal case provide a check on the multimode cases we consider later in the paper, as well as providing a bound on the best possible squeezing for the system.

In Section \ref{sec:MMdescription} we show how multimode behaviour becomes important in physical systems, and introduce our model for simulating higher dimensional trap geometries.  Section~\ref{sec:1Dsim} analyses one-dimensional models, highlighting some of the issues regarding optimising squeezing parameters and mode-matching in squeezing experiments.  Section~\ref{sec:pipulse} shows that the application of a $\pi$-pulse during the process can significantly improve mode-matching issues.  Section~\ref{sec:2D3Dsqueezing} investigates measurable squeezing in 2D and 3D, and uses a Bogoliubov analysis to explain the scaling with dimensionality. Section~\ref{sectionConclusion} describes the conclusion that large volume, flat-bottomed, three-dimensional traps, which are obviously compatible with large atom number, can produce extremely high degrees of squeezing.

@@

When considering squeezing, the Hamiltonian \ref{eqTwoModeHamiltonian} has typically been handled within the framework of spin squeezing, which has a long history \cite{kitagawaET1993, esteve2008,sorensenET1999, liebfriedET2004, wineland1994, steel+collett}. This approach involves exploiting the equivalence between the algebra of two harmonic oscillators and that of angular momentum, and writing
\begin{eqnarray}
\hat{S}_+ &=& \hat{a}^{\dagger} \hat{b} \\
\hat{S}_- &=& \hat{a} \hat{b}^{\dagger} \\
\hat{S}_x &=& \frac{1}{2} \left( \hat{S}_+ + \hat{S}_-\right) \\
\hat{S}_y &=& \frac{1}{2i} \left( \hat{S}_+ - \hat{S}_-\right) \\
\hat{S}_z &=& \frac{1}{2} \left( \hat{a}^{\dagger} \hat{a} - \hat{b}^{\dagger} \hat{b} \right).
\end{eqnarray}
Using these variables Eq.~(\ref{eqTwoModeHamiltonian}) is transformed into the Josephson Hamiltonian \cite{steel+collett}
\begin{equation}
\Delta \omega \, \hat{S}_z + \chi \hat{J}_z^2 + \Omega \hat{J}_x
\label{eqJosephson}
\end{equation}
where the $\Delta \omega$ is a function of the various nonlinearities and $\chi=\chi_{aa}+\chi_{bb}-2\chi{ab}$ are functions of the various nonlinearities, which is then solved using angular momentum algebra.

We take a different approach and solve Eq.~(\ref{eqTwoModeHamiltonian}) directly for two reasons. First, we want access to the number variances of each state, not just the variance of the number difference between the states. Second, Eq.~(\ref{eqJosephson}) has led to the common misconception that no squeezing can be generated in this system if $\chi_{aa}+\chi_{bb}-2\chi{ab}=0$, as the effects of the nonlinearities on $\Delta \omega$ are dismissed as an irrelevent detuning. This is only the case if a spin-echo pulse is used half way through the application of (\ref{eqTwoModeHamiltonian}) (which is the protocol used in most spin squeezing experiments) or extra detunings are added to (\ref{eqJosephson}). Directly solving the Hamiltonian (\ref{eqTwoModeHamiltonian}) makes this clearer.

The system is prepared with all the population initially in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to initially be a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As the system is insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states. Such a state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

@@

We solve this system analytically, and derive expressions for absolute and relative number squeezing. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation} as it is somewhat laborious. If we define
\begin{eqnarray}
s &=& \sin(\theta) \label{eqsDef} \\
c &=& \cos(\theta) \\
\theta &=& \Omega(t_3 - t_2) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \nonumber\\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) \left( n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B \right. \nonumber \\
       && \,\,\,\,\, \left. + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^* \right) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{johnssonET2007,haineET2009,steelET1998,dennisET2010,johnssonHope:2007}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against both the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such as the appearance of negative densities in lightly populated modes.  We could also use the Bogoliubov analysis described in section \ref{sec:2D3Dsqueezing} as an independent estimate of the validity of the TWA.  We found that the TWA remained valid except in some 3D simulations with high nonlinearities, where there were a large number of modes, and the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown.  Fortunately, the best squeezing was typically found in regions where the TWA was valid.

@@

We have presented an analytic solution to the two-mode model of atomic fields with a strong Kerr non-linearity, allowing for a rapid estimate of the atomic number difference squeezing that can be obtained from these systems as scattering length data is obtained. In general, it predicts arbitrarily good squeezing in the two-mode limit. We then examined trap geometries in which this limit can be approached.

Using a constant potential in the trap (a `box' trap) allows the BEC density to be spatially uniform throughout the process.  This is shown to alleviate the difficulties with choosing correct hold times and phases for the coupling pulses, and also alleviates the majority of the mode-matching issues for the final recombination of the two modes. If such a box potential is not available, the application of a $\pi$-pulse part way through the hold period can also improve mode matching by imparting the same modal dynamics to both states.

@@

\bibitem{Leggett2001} Leggett A J 2001 Bose-Einstein condensation in the alkali gases: Some fundamental concepts \emph{Rev. Mod. Phys.} {\bf 73} 307

\bibitem{Buluta2009} Buluta I and Nori F 2009 Quantum simulators \emph{Science} {\bf 326} 108

\bibitem{Kinoshita:2006} Kinoshita T, Wenger T R and Weiss D S 2006 A quantum Newton's cradle \emph{Nature} {\bf 440} 900

\bibitem{Barnett:2011} Barnett R, Polkovnikov A and Vengalattore M 2011 Prethermalization in quenched spinor condensates \emph{Phys. Rev.} A {\bf 84} 023606

@@

\bibitem{savage2007} Savage C M and Kheruntsyan K V 2007 Spatial pair correlations of atoms in molecular dissociation \emph{Phys. Rev. Lett.} {\bf 99} 220404

\bibitem{jascula2010} Jaskula J C, Bonneau M, Partridge G B, Krachmalnicoff V, Deuar P, Kheruntsyan K V, Aspect A, Boiron D, and Westbrook C I 2010 Sub-Poissonian number differences in four-wave mixing of matter waves \emph{Phys. Rev. Lett.} {\bf 105} 190402

\bibitem{esteve2008} Est{\`{e}}ve J, Gross C, Weller A, Giovanazzi S, and Oberthaler M 2008 Squeezing and entanglement in a Bose-Einstein condensate \emph{Nature} {\bf 455} 1216

\bibitem{pu2000} Pu H and Meystre P 2000 Creating macroscopic atomic Einstein-Podolsky-Rosen states from Bose-Einstein condensates \emph{Phys. Rev. Lett.} {\bf 85} 3987

@@

@@

\bibitem{wineland1994} Wineland D J, Bollinger J J, Itano W M and Heinzen D J 1994 Squeezed atomic states and projection noise in spectroscopy \emph{Phys. Rev.} A {\bf 50} 67

\bibitem{steel+collett} Steel M and Collett M 1998 Quantum state of two trapped Bose-Einstein condensates with a Josephson coupling \emph{Phys. Rev.} A {\bf 57} 2920

@@

\bibitem{gaunt2013} Gaunt A L, Schmidutz T F, Gotlibovych I, Smith R P, and Hadzibabic Z 2013 Bose-Einstein condensation of atoms in a uniform potential \emph{Phys. Rev. Lett.} {\bf 110} 200406

\bibitem{cornish:2000} Cornish S, Claussen N, Roberts J, Cornell E, and Wieman C 2000 Stable $^{85}$Rb Bose-Einstein condensates with widely tunable Interactions \emph{Phys. Rev. Lett.} {\bf 85} 1795

@@
